Large-scale spatiotemporal variation in vital rates of an alpine bird

### 2023-06-06 Manuscript_Nater_etal.qmd
Added some worknote content

---
title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
format: html
author: Chloé R. Nater, James Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

## Introduction

## Methods

## Results

## Discussion

### Conclusion

### Author contributions

### Acknowledgements

### Data and code availability

>>>>

---
title: "Large-scale spatiotemporal variation in vital rates of an alpine bird"
format: html
author: Chloé R. Nater, James A. Martin, Erlend B. Nilsen
editor: visual
bibliography: rypeIDSM_MS2.bib
---

## About this worknote
The purpose of this worknote is to introduce and showcase preliminary results from application of the open-population integrated distance-sampling model (IDSM) for willow ptarmigan to data from all of Norway. Eventually, this model extension and resulting insights into large-scale spatiotemporal variation in density, population growth rate, and vital rates will be the subject of a second scientific paper. This paper will be written and submitted in the second half of 2023, and this worknote serves as a temporary substitute for it in the reporting of project results to Miljødirektoratet in June 2022. 
As such, the worknote contains a general description of the extension of the model to analyse data from multiple areas simultaneously, visualizations of some preliminary results, and an introduction to a reproducible workflow that allows to easily and robustly re-run analyses as new data becomes available in future years.

## IDSM extension to multiple areas

Distance-sampling data on willow ptarmigan is collected on line transects not only in western Lierne, but across most of its Norwegian range (https://honsefugl.nina.no/). Taking the line transect datasets from Fjellstyrene, Statskog, and FeFo together, there is data for willow ptarmigan in 41 reporting areas (for a list, see https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp/blob/main/R/listAreas.R). 
The model developed as part of this project and presented in @nilsen2023 was designed with the potential of extension to run on data from multiple areas simultaneously in mind, and this extension involves two steps in practice.

The first step is the inclusion of an additional dimension, area, to enable using data collected from multiple areas (@fig-model). The area-dimension is added across all model components with the exception of the known-fate telemetry data (and seasonal decomposition of survival estimated from it), as that is only available for the Lierne area.

![Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model. The pink "x"'s indicate the added dimension for area. Juv\[x,t\] = juveniles in year t. Ad\[x,t\] = adults in year t. R\[x,t\] = recruitment rate in year t. S\[x,t\] = survival probability from year t to t+1.](Figures/RypeIDSM_multiArea_structure.png){#fig-model}

The second step is making adjustments to capitalize on the integration of data from multiple areas to make maximal use of available information. 
There are several not mutually exclusive options for approaching this, and one of them is the use of "hyper-parameters" for sharing information across areas [e.g. @horswill2019; @horswill2021].
As an illustrative example, take the area-specific average annual survival probability, $\mu^S_x$. Instead of assigning independent non-informative priors to each $\mu^S_x$, we now assume that the variation in average survival across space can be described by a common (logit) normal distribution such that:

$$
\begin{equation}
\mu^S_x \sim Normal(logit(h.\mu^S), h.\sigma^S)
\end{equation}
$$
where $h.\mu^S$ is the "hyper mean" and $h.\sigma^S$ is the "hyper standard deviation". 
Non-informative priors are then assigned to the hyper-parameters instead of the area-specific realizations:

$$
\begin{align}
h.\mu^S & \sim Uniform(0,1) \\
h.\sigma^S & \sim Uniform(0, 5)
\end{align}
$$

For the multi-area implementation of the IDSM, we used the equivalent of the above approach for parameters representing average survival probabilities, baseline recruitment rates, rodent effect on recruitment, and average distance sampling detection.
Initial population density, on the other hand, was estimated independently for each area. 
Random effects standard deviations were assumed to be the same for all areas, effectively resulting in space-for-time subtitution for random effect estimation.

Another approach to maximising information gain when jointly analysing data from different areas is explicit modelling of spatial and spatio-temporal variation. While this is commonly done for modern species distribution models [e.g. @pacifici2017; @guelat2018], it has rarely been implemented in demographic models. This is not least due to the fact that few demographic models span large spatial extents. Since our IDSM and the willow ptarmigan data do actually do that though, extending the framework to account for spatial and spatiotemporal correlations may be feasible. Per now, we have not yet attempted this, but are planning to do so in the near future.

## Preliminary results

### Population density

::: {#fig-dens layout-ncol=2}

![Median density estimates](Figures/density_2018-2020_Median.png){#fig-densA}

![Uncertainty in density estimates](Figures/density_2018-2020_SD.png){#fig-densB}

Median (a) and standard deviation (b) of posterior estimates of average ptarmigan density in the three most recent years (2018-2020) across 41 reporting areas in Norway.
:::

### Population growth rate

::: {#fig-lam layout-ncol=2}

![Median population growth rate estimates](Figures/lambda_2018-2020_Median.png){#fig-lamA}

![Uncertainty in population growth rate estimates](Figures/lambda_2018-2020_SD.png){#fig-lamB}

Median (a) and standard deviation (b) of posterior estimates of average annual population growth rate over the three most recent years (2018-2020) across 41 reporting areas in Norway.
:::

### Survival probability

::: {#fig-pSurv layout-ncol=2}

![Median survival estimates](Figures/pSurv_Median.png){#fig-pSurvA}

![Uncertainty in survival estimates](Figures/pSurv_SD.png){#fig-pSurvB}

Median (a) and standard deviation (b) of posterior estimates of average annual survival probability across 41 reporting areas in Norway.
:::

### Recruitment rate

::: {#fig-rRep layout-ncol=2}

![Median recruitment estimates](Figures/rRep_Median.png){#fig-rRepA}

![Uncertainty in recruitment estimates](Figures/rRep_SD.png){#fig-rRepB}

Median (a) and standard deviation (b) of posterior estimates of average annual recruitment rate across 41 reporting areas in Norway.
:::

### Effect of rodent occupancy

::: {#fig-betaR layout-ncol=2}

![Median rodent effect estimates](Figures/betaR_Median.png){#fig-betaRA}

![Uncertainty in rodent effect estimates](Figures/betaR_SD.png){#fig-betaRB}

Median (a) and standard deviation (b) of posterior estimates of effects of rodent occupancy on recruitment rate across 41 reporting areas in Norway.
:::

## Reproducible workflow with "targets"

## Data and code availability
